MWA tied-array processing III: Microsecond time resolution via a synthesis filter

### 2019-10-18 ipfb.tex
Finished first draft of the PFB description section

Here, we describe the synthesis filter that is implemented as the (optional) final stage of the tied-array beamforming pipeline, the former stages of which are described in detail in \citet[][hereafter Paper I]{Ord2019} and \citet[][hereafter Paper II]{Xue2019}.
The synthesis filter is applied to the fine-channel output of the beamformer, and recovers the coarse channel time series.
That is, it effectively `undoes' the fine PFB, increasing the available time resolution to $\sim 0.8\,\mu$s.
A brief review of PFBs in general, and their particular implementation in the case of the MWA, are given in \S\ref{sec:pfb}.
The design of the synthesis filter is described in \S\ref{sec:ipfb}, including a discussion of its fidelity, i.e., the appearance of any temporal and spectral artefacts introduced by the synthesis filter itself.
Finally, the practical use of this functionality is demonstrated through two examples: observations of the MSPs J2241$-$5236 and J0437$-$4715.

@@

PFBs are a type of analysis filter, designed to extract spectral information out of discrete time series data.
They can be considered a generalisation of the more familiar Discrete Fourier Transform (DFT), and are designed to overcome the dual problems of uneven frequency response and spectral leakage inherent in the application of DFTs to finite length data sets.
They are well described in standard texts \citep{Crochiere1983,Harris2004,Oppenheim2009}; a clear and concise review of PFBs in the context of radio astronomy is given in \citet{Harris2011}.
Therefore, only a brief review of the salient features are given below, in order to lay out the notation used in this paper and to prepare the reader for what follows.

\subsection{General review and mathematical description}

>>>>

Here, we describe the synthesis filter that is implemented as the (optional) final stage of the tied-array beamforming pipeline, the former stages of which are described in detail in \citet[][hereafter Paper I]{Ord2019} and \citet[][hereafter Paper II]{Xue2019}.
The synthesis filter is applied to the fine-channel output of the beamformer, and recovers the coarse channel time series.
That is, it effectively `undoes' the fine PFB, increasing the available time resolution to $\sim 0.8\,\mu$s.
A brief review of PFBs in general, and their particular implementation in the case of the MWA, are given in \S\ref{sec:pfb}.
The design of the synthesis filter is described in \S\ref{sec:ipfb}, including a discussion of its fidelity, i.e., the appearance of any temporal and spectral artefacts introduced by the synthesis filter itself.
Finally, the practical use of this functionality is demonstrated through two examples: MWA observations of the MSPs J2241$-$5236 and J0437$-$4715.

@@

PFBs are a type of analysis filter, designed to extract spectral information out of discrete time series data.
They can be considered a generalisation of the more familiar discrete Fourier transform (DFT), and are designed to overcome the undesirably uneven frequency response (i.e. spectral leakage) inherent in the application of DFTs to discretely sampled time series of finite length.
They are well described in standard texts \citep{Crochiere1983,Harris2004,Oppenheim2009}; a clear and concise review of PFBs in the context of radio astronomy is given in \citet{Harris2011}, whose notation is closely followed here.
Thus, only a brief review of the salient features is given, in order to prepare the reader for the description of the synthesis filter in the following sections.

\subsection{General review and mathematical notation}

Like the DFT, the PFB is a transformation from the time domain, $x[n]$, to the frequency domain $X[k]$, where the $[\cdot]$ notation denotes a discretised index, and where $n,k \in \mathbb{Z}$ are the indices for time and frequency, respectively.
Let $N$ be the number of (equally spaced) frequency channels required for some desired spectral resolution.
Although applying a DFT to $N$ adjacent time samples will produce the desired resolution, the result will be an imperfect representation of the true spectral content of $x[n]$ because of \textit{spectral leakage}, which is when power that properly `belongs' to some particular frequency bin appears in (or, is \textit{aliased} to) other, nearby bins.
The impossibility of perfectly eliminating spectral leakage can be seen by realising that operating on a finite-length time series is equivalent to multiplying an arbitrarily long time series with a rectangular window function,
\begin{equation}
    w_R[n] = \begin{cases} 1, & 0 \le n < N, \\ 0, & \text{elsewhere}, \end{cases}
\end{equation}
whose effect in the frequency domain is to convolve the ``true'' spectrum of the signal with the Fourier transform of the window function.
In the case of a rectangular window, this is the sinc function.

A common strategy for mitigating spectral leakage is to choose an alternative window function whose Fourier pair is localised in the frequency domain, and which therefore produces a tolerable level of spectral leakage when convolved with the signal's spectrum.
Many possible windowing functions have been identified, which in general trade the `amount' of leakage with the `location' of the leaked components.
For our purposes, it is sufficient to note that the inevitable presence of a windowing function motivates the definition of the \textit{analysis filter}, $h[n]$, and the generalised \textit{windowed DFT}
\begin{equation}
    X[k] = \sum_{n=0}^{N-1} h[n]\,x[n]\,e^{-2\pi j k (n/N)},
    \label{eqn:wdft}
\end{equation}
where $j = \sqrt{-1}$ denotes the imaginary number, and it is understood that $0 \le k < N$.
The choice of $h[n]$ is motivated by the shape of its \textit{frequency response} (i.e. its Fourier pair), whose characteristics (e.g. width, location of sidelobes) might carry particular advantages in certain contexts.
Note that leaving $x[n]$ ``unweighted'' is equivalent to choosing $h[n] = w_R[n]$.

It is well known that scaling a function in the time domain produces the inverse scaling in the Fourier domain.
This fact motivates an alternative strategy for mitigating spectral leakage.
If more samples are available, say $M = NP$ for $P > 1$, then stretching (i.e. scaling and upsampling) the analysis filter will result in a frequency response function that is the same shape as, but $P$ times narrower than the original analysis filter.
Applying the DFT to the larger set of windowed samples and subsequently \emph{downsampling} the resulting spectrum (by choosing every $P$th frequency bin and rejecting the rest) ensures that the original desired spectral resolution is retained, but with significantly less spectral leakage.

The algorithm embodied in the above strategy defines the PFB.
It is clear that choosing a larger value for $P$ will lessen spectral leakage by the same factor, but this naturally comes at a greater computational cost.
However, one of the great advantages of the PFB is that the above operations of using a wider analysis wilter and downsampling the resulting spectrum are mathematically equivalent to first segmenting the windowed time series into segments (known as `taps') of size $N$, summing their respective samples element-wise, and performing a single DFT on the resulting array (still of size $N$).
Mathematically,
\begin{equation}
    X[k] = \sum_{n=0}^{N-1} b[n]\,e^{-2\pi j k (n/N)},
    \label{eqn:pfb}
\end{equation}
where
\begin{equation*}
    b[n] = \sum_{\rho=0}^{P-1} h[n + N\rho]\,x[n + N\rho].
\end{equation*}
A short proof of the equivalence of this expression with the algorithm described above is given in \citet{Harris2011}.
The name `polyphase' refers to the structure of $b[n]$ defined in Eq. \eqref{eqn:pfb}.
It is illustrated in Fig. \ref{fig:pfb}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{example-image-a}
    \caption{A diagramatic representation of the PFB algorithm, as defined in Eq. \eqref{eqn:pfb}.}
    \label{fig:pfb}
\end{figure}
